Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
add code for memo combinators, use smallverbatim

In this second approach, we used the
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array).
We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{Durbin:1998wz} or the HMMER literature~\cite{Eddy:1998ut} 
(see Equation \ref{viterbi_eqn}).
We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases.
This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node.
In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down 
dynamic-programming approach in Haskell, which resulted in perspicuous code.

@@

We wrote a basic search function in only twenty-two lines of code.
The central logic of this search function is simply:
\begin{verbatim}
search' (s1:s2:seeds) oldPop hist age =
  let newPop = mutate oldPop
      score = fst $ minimum newPop
      newHist = score : hist
    in if accept newHist age then
          if terminate newHist age then
            (minimum newPop, newHist)
          else
            search' seeds newPop newHist (age + 1)
       else
         if terminate hist age then
           (minimum oldPop, hist)
         else
           search' seeds oldPop hist (age + 1)
\end{verbatim}
As one of the parameters to the top-level \texttt{search} function,
we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques.
We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies.
For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions.
In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}.
We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.
In an imperative language with object orientation like Ruby 
or Python, we likely would have used a class hierarchy to accomplish this.
We 
feel that this would have required more code, because object orientation is 
meant for encapsulating data as well as functions on those data; here we only 
wanted to encapsulate functions, as the data did not change.

>>>>

In this second approach, we used the
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array).
This combinator approach was easy to apply to the code after the fact:
\begin{smallverbatim}
viterbi state node obs = 
  Memo.memo3 (Memo.arrayRange (mat, end)) 
             (Memo.arrayRange (0, numNodes))
             (Memo.arrayRange (0, seqlen)) 
             viterbi' state node obs
\end{smallverbatim}
We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{Durbin:1998wz} or the HMMER literature~\cite{Eddy:1998ut} 
(see Equation \ref{viterbi_eqn}).
We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases.
This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node.
In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down 
dynamic-programming approach in Haskell, which resulted in perspicuous code.

@@

We wrote a basic search function in only twenty-two lines of code.
The central logic of this search function is simply:
\begin{smallverbatim}
search' (s1:s2:seeds) oldPop hist age =
  let newPop = mutate oldPop
      score = fst $ minimum newPop
      newHist = score : hist
    in if accept newHist age then
          if terminate newHist age then
            (minimum newPop, newHist)
          else
            search' seeds newPop newHist (age + 1)
       else
         if terminate hist age then
           (minimum oldPop, hist)
         else
           search' seeds oldPop hist (age + 1)
\end{smallverbatim}
As one of the parameters to the top-level \texttt{search} function,
we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques.
We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies.
For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions.
In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}.
We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.
In an imperative language with object orientation like Ruby 
or Python, we likely would have used a class hierarchy to accomplish this.
We 
feel that this would have required more code, because object orientation is 
meant for encapsulating data as well as functions on those data; here we only 
wanted to encapsulate functions, as the data did not change.
